Temporal ordering and registration of imaging data in cross--sectional studies of development

### 2014-04-09 image_analysis_paper/pnas_draft.tex
updated draft with some of Yannis comments

\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim\affil{1}{}, 
Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Hang~Lu\affil{2}{},
Stanislav~Y.~Shvartsman\affil{1}{} \affil{3}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}, 
\and 
Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}

@@

\begin{abstract} 
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time. 
%
The goal is then to temporally order the data to reconstruct the developmental dynamics. 
%
If each data point is a two-dimensional image, the images must first be aligned or registered before they can be temporally ordered. 
%
When such data sets are large, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand. 
%
We present an automatic way to register and temporally order cross--sectional data sets of images. 
%
We use mathematical techniques (angular synchronization for image registration, and diffusion maps for temporal ordering) which are applicable to a wide variety of data sets and require little {\em a priori} knowledge of the image features or the developmental dynamics. 
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional. 
%
In longitudinal studies, developmental progress is monitored over time within the same embryo. 
%
In a cross-sectional study, one embryo contributes only one snapshot of a chemical or morphological process along a developmental trajectory, which must be reconstructed from multiple snapshots of different embryos. 
%
Both of these experimental designs have their advantages and limitations and are extensively used by developmental biologists. 
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms. 
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes. 
%
Fixed embryos are then imaged using any given number of microscopic techniques. 
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different orientation and time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy. 
%
In general, it is known that a collection of embryos belongs to a certain time window.
%
In order to mine such datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned and ordered in time. 
%
We present a general framework that greatly accelerates both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are apparent. 
%
As an example, Figure~\ref{fig:fish} shows a schematic of fish development, combining the processes of growth and patterning.  
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which correlates with the progress of development, and image registration is based on obvious morphological landmarks, such as positions of head and fins. 
%
Dealing with real data is far more nontrivial, due to the number of images and similarities between them. 
%
To illustrate most of the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of pattern formation in the early fruit fly embryo, one of the leading systems for studies of developmental dynamics. 
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now registered and temporally ordered by hand. For this example, the registration and ordering is easy to do by hand because the data set is small and the developmental changes are easy to recognize.} 
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long. 
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged in under the common plasma membrane. 
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in length (Figure~\ref{subfig:membranes}). 
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}. 
%
Images in Figure~\ref{subfig:images_unordered} show optical sections through the short axis of the embryos. 
%
The embryos were fixed and stained with chemicals that visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}. 
%
Each of the images reveals a uniform arrangement of nuclei and nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumfrence of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time. 
%
The goal is to align these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory. 
%
Below we show that both of these tasks can be fully automated.
%
We will use angular synchronization \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use diffusion maps \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using vector diffusion maps \cite{singer2012vector}.
%
%The paper is organized as follows. 
%%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering. 
%%
%Then, we show that both of these tasks can be effectively combined in a single step. 
%%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time. 
%%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%%
%However, this algorithm requires distances between data points. 
%%
%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
%%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is in an arbitrary orientation; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discritize the profiles into 100 points, so our data will be 100-dimensional vectors. 
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring. 
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set. 
%
The concentration profiles have been stacked in an array, such that each row corresponds to one profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian growing in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) Many concentration profiles, each of the form described in (a). Each row in the array corresponds to one image. (d) The concentration profiles in (c), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. (f) The concentration profiles in (c), now simultaneously registered and temporally ordered using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

Our first task is to register the set of images or profiles.
%
%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}. 
%
We will use angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first computes the pairwise alignments %(rotations for our one-dimensional profiles in Figure \ref{fig:1d_demo}c, or translations and rotations for the two-dimensional images in Figure \ref{fig:images_unordered}) 
for every possible pair of data points. %, effectively using each image as a template for every other image.
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
%
From this large collection of pairwise alignments, angular synchronization then rotates each data point, such that the individual alignments are most consistent (in some metric) with the pairwise alignments.
% (i.e., we would like the difference between the optimal rotations for two different concentration profiles to be as close as possible to the pairwise rotation needed to align the two profiles).
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned.
%-- these are the angles we would like to recover from the pairwise information. 
%
Note that the differences between the rotation angles are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
To compute these alignments, we formulate the task as an optimization problem, and then relax it to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about the pairwise alignments) give the alignments for each data point.
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (b) The vectors from (a), each rotated so that the set of vectors is aligned. Note that the angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge intensity is related to the distance between the points (close data points are connected by dark edges). (d) The data in (c), colored by the first (non-trivial) eigenvector from diffusion maps, $\phi_2$. The color parameterizes the one-dimensional curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
The peaks are now aligned, even though our algorithm use no information about the existence or location of these peaks.
%
Because angular synchronization only uses pairwise rotation information, the optimal rotations are recovered up to a global rotation.

@@

After registering the images or concentration profiles, we want to order them in time to reconstuct the developmental dynamics. 
%
We will use diffusion maps (DMAPS) to temporally order our data.
%
In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in a high-dimensional data set \cite{coifman2005geometric}. 
%
Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional data'') which lies on a one-dimensional nonlinear curve (the low-dimensional structure).
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the first diffusion maps coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect.
%
We assume that our imaging data also falls on a one-dimensional, nonlinear curve, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve). 
%
Therefore, ordering data along this curve will effectively order data in time.
%
Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.

To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close on the curve (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have a dark edge between them.
%
Visually, the dark edges are meaningful and delineate the one-dimensional curve.
%
Furthermore, the light edges are not informative and do not depict any meaningful structure. 
%
We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional parameterization.
%
This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a matrix which contains information about pairwise distances between the data points. 
%
In general, this procedure is called diffusion maps \cite{coifman2005geometric}.%; the formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
%
We use the Euclidean distance between the aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$. 
%
As expected, the intensity of the peak grows as a function of time. 
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

\subsection{Vector diffusion maps for registration and ordering}

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
We can combine angular synchronization (for registration) and diffusion maps (for temporal ordering) into one eigencomputation that allows us to {\em simultaneously} recover the optimal alignments and the temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector}.
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps. 
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate.

>>>>

\author{Carmeline~J.~Dsilva\affil{1}{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim\affil{1}{},
Thomas~J.~Levario\affil{2}{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Hang~Lu\affil{2}{},
Amit~Singer\affil{3}{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA},
Stanislav~Y.~Shvartsman\affil{1}{} \affil{5}{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA},
\and
Ioannis~G.~Kevrekidis\affil{1}{} \affil{4}{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}}

@@

\begin{abstract}
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a slightly different developmental time.
%
The goal is then to temporally order the data to reconstruct the developmental dynamics.
%
If each data point is a two-dimensional image, the images must first be registered (aligned) before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to do by hand.
%
We present an automatic approach to registering {\em and} temporally order cross--sectional data sets of images.
%
The mathematical techniques (angular synchronization for image registration, diffusion maps for temporal ordering and
vector diffusion maps for simultaneously performing both tasks) are applicable to a wide variety of data sets and 
require little {\em a priori} knowledge of the image features or the developmental dynamics.
%
We demonstrate the utility of our methods using a collection of images from a study of {\em Drosophila} embryogenesis.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal \cite{...} and cross-sectional \cite{...}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo.
%
In a cross-sectional study, one embryo contributes only a single snapshot of a chemical or morphological process along its developmental trajectory; the latter must be reconstructed from multiple snapshots of different embryos.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by developmental biologists.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of developing embryos is fixed using a procedure that arrests their development and stained with chemicals that help visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale physical manipulation and imaging of embryos have produced rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different geometric orientation and developmental time point.
%
Importantly, the ``age'' of any given embryo arrested in its development is not known to high accuracy.
%
In general, it is only known that a collection of embryos belongs to a certain time window.
%
In order to mine such image datasets for underlying developmental dynamics, snapshots of different embryos must be spatially aligned ("registered")  and ordered in time.
%
We present a general computational framework that greatly accelerates -and can combine- both of these tasks.

Temporal ordering and registration of images is straightforward when the number of images is small and differences between them are visually apparent.
%
As an example, Figure~\ref{fig:fish} shows a caricature of fish development, combining the processes of growth and color patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which uniquely correlates with the progress of development; image registration is then based on obvious morphological landmarks, such as positions of head and fins.
%
Real data will, of course, provide nontrivial challenges, due to the number of images, due to variability and measurement noise, and due to the subtlety of several developmental changes.

%
To illustrate the issues associated with temporal ordering and registration of images, we use a data set from an imaging study of regulatory protein expression 
%%%
%%% check whether "regulatory protein" is the correct term
%%%
pattern formation in the early fruit fly embryo, one of the leading model systems for studies of developmental dynamics.
%
This dataset comprises $49$ fluorescence microscopy images of different embryos fixed during the $3^{rd}$ hour of their development.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig1}
\caption{Schematic illustration of reconstructing dynamics from cross-sectional data. (a) Fish, each in a different orientation and a different stage of development. (b) Fish, now rotationally registered and temporally ordered. For this example, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}a}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}b}
\end{figure}

The fruit fly embryo is shaped like a rice grain, approximately $1/2$~mm long.
%
By end of the $2^{nd}$ hour of development, sequential divisions of a fertilized egg generate a system where $\sim 6,000$ nuclei are uniformly arranged under the common plasma membrane.
%
During the $3^{rd}$ hour, the embryo undergoes the process of cellularization, whereby membranes that enclose nuclei into different cells grow in thickness (Figure~\ref{subfig:membranes}).
%
Concurrently with these morphological changes, concentration profiles of multiple regulatory molecules form inside the embryo, subdividing it into regions that give rise to future tissues and organs \cite{lim2013kinetics}.
%
Images in Figure~\ref{subfig:images_unordered} show optical sections perpendicular to the long axis of the embryo.
%%%
%%% we have to somehow say that it is consistent
%%%
%
%%%
%%% ask if "visualize" is typical as active verb
%%%
The embryos were fixed and stained with chemicals that help visualize nuclei (shown in gray) as well as the spatial distributions of two different proteins (shown in green and red) \cite{chung2010microfluidic}.
%
Each of the images reveals a uniform arrangement of nuclei and an azimuthally nonuniform distribution of the two proteins, which are involved in patterning of the dorsoventral (back-to-belly) axis of the embryo.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig2}
\caption{(a) Fluorescent images of {\em Drosophila} embryos. Each embryo has been stained for Dorsal (green) and dpERK (red). (b) Selected image from data set. (c) Profile of Dorsal (green) and dpERK (red) around the circumference of the embryo in (b).}
%\label{fig:images_unordered}
\customlabel{fig:images_unordered}{2}
\customlabel{subfig:images_unordered}{\ref{fig:images_unordered}a}
\customlabel{subfig:one_image}{\ref{fig:images_unordered}b}
\customlabel{subfig:image_profiles}{\ref{fig:images_unordered}c}
\end{figure}

%%% 
%%% orientation does not read good to Y below -
%%% it is the same orientation, but a different rotation around the axis
%%% how to say this ?
%%%
Each image in Figure~\ref{subfig:images_unordered} comes from a different embryo, captured at a different orientation and at a different and unknown point in time.
%
The goal is to register these images in a globally consistent way and order them in a way that is most representative of the underlying developmental trajectory.
%
Both of these tasks can be fully automated, and even combined.
%
We will use {\em angular synchronization} \cite{singer2011angular}, a registration algorithm that is robust to noise, to align the images in a consistent way.
%
We will use {\em diffusion maps} \cite{coifman2005geometric}, a broadly applicable technique for the discovery of low-dimensional structures in high-dimensional data sets, to temporally order our data.
%
We will also show how both steps can be combined using {\em vector diffusion maps} \cite{singer2012vector}.
%
%The paper is organized as follows.
%
%First, we use a synthetic dataset, where the correct orientation and ordering of data points is known, to formalize the tasks of image registration and temporal ordering.
%
%Then, we show that both of these tasks can be effectively combined in a single step.
%
%Following this, we return to the original set of two-dimensional images and demonstrate that our approach can readily align them and order them correctly in time.
%
%We conclude by discussing the general requirements that must be satisfied for our approach to work and compare it to other approaches and ideas in cross-sectional studies of biological dynamics.

@@

%We would like to temporally order images, such as the images shown in Figure~\ref{subfig:images_unordered}.
%
%We will use diffusion maps, a dimensionality reduction algorithm, to automatically order our data.
%
%However, this algorithm requires distances between data points.
%
%Because each image is in a different orientation, we need to first align or {\em register} the images before computing distances.
%
%Once all the images have been aligned, we can compute distances and use diffusion maps to temporally order the data.

%%%
%%%  Y to discuss with S
%%%  I think we should have the real data
%%%
%%%  also, these dynamics "are like" our dynamics
%%%
%%%   even if this is kept, one should say that this chosen because it LOOKS LIKE what we have
%%% another reason I want to have the real data is because in the discussion I want to talk about the flips 
%%%

We will first demonstrate our techniques for registration and temporal ordering using a synthetic data set with relatively simple dynamics.
%
We construct concentration profiles defined on a ring, and we assume that each ring is arbitrarily rotated around its center; an example is shown in Figure~\ref{subfig:1d_example}.
%
Each concentration profile is a noisy Gaussian (shown in Figure~\ref{subfig:1d_intensity}), and the Gaussians increase in intensity as  a function of time.
%
We discretize the profiles into $100$ points, so our numerical data will be $100$-dimensional vectors.
%
Rotation of the ring in Figure~\ref{subfig:1d_example} corresponds to shifting (with periodic boundary conditions) the one-dimensional concentration profile shown below the ring (the continuous symmetry group is the group of all two-dimensional proper rotations, denoted $SO(2)$; the corresponding discrete symmetry group for the discretized profiles is $\mathbb{Z}_n$, the group of integers modulo $n$).
%
Figure \ref{subfig:1d_unaligned_unordered} shows all the concentration profiles in our data set; they have been stacked in an array, so that each row corresponds to a single profile.
%
Because the profiles are unregistered and unordered, the underlying dynamics (a Gaussian whose amplitude grows in time) are not readily apparent.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig3}
\caption{Illustration of methodologies using one-dimensional profiles. (a) One-dimensional concentration profile on a ring (top), and the corresponding profile on a line (bottom). (b) Intensity corresponding to the profile in (a). (c) An ensemble of concentration profiles, each of the form described in (a). Each row in the array corresponds to a single image. (d) The concentration profiles in (c), now registered using angular synchronization. (e) The concentration profiles in (c), now temporally ordered using diffusion maps. (f) The concentration profiles in (c), registered and temporally ordered in a single step using vector diffusion maps.}
%\label{fig:1d_demo}
\customlabel{fig:1d_demo}{3}
\customlabel{subfig:1d_example}{\ref{fig:1d_demo}a}
\customlabel{subfig:1d_intensity}{\ref{fig:1d_demo}b}
\customlabel{subfig:1d_unaligned_unordered}{\ref{fig:1d_demo}c}
\customlabel{subfig:1d_aligned_unordered}{\ref{fig:1d_demo}d}
\customlabel{subfig:1d_aligned_ordered}{\ref{fig:1d_demo}e}
\customlabel{subfig:1d_aligned_ordered_vdm}{\ref{fig:1d_demo}f}
\end{figure}

@@

We start by registering the set of images or profiles.
%
%``Template-based'' registration \cite{ahuja2007template} is a common technique, where one selects a specific image or {\em template}, and then translates and rotates each image in the data set until it is optimally aligned with this template (optimal for us means minimizing the norm between the difference in pixel intensities).
%
%However, this method can result in many errors if the images are noisy, and can be problematic if a ``good'' template image is not known {\em a priori}.
%
We will use angular synchronization\cite{singer2011angular}, a technique which is robust to noise and requires no {\em a priori} knowledge of image features.
%
Angular synchronization first aligns every possible pair of data points; that is, it computes the rotation angle
required to make them have the maximum possible overlap (see {\it Materials and Methods}).
%%%
%%% so I do not like the figure with the vectors - it is correct, but people have to take the
%%% function, discretize and make it a vector in their heads, and may be "one step too much"
%%%
%
An illustration is shown in Figure~\ref{subfig:synch1};
each data point is a vector, and $\theta_{ij}$ denotes the angle of rotation needed to align vectors $i$ and $j$ (these are the pairwise alignments).
%
From this large collection of pairwise alignments, angular synchronization then selects a unique (up to a global rotation)
set of angles $\theta_i$, one for each data point, such that the resulting pairwise alignments $\theta_i-\theta_j$ are {\em most consistent} in some metric.
%
Figure~\ref{subfig:synch2} shows each vector from Figure~\ref{subfig:synch1}, now rotated so that the entire set is aligned
(the rotation angles we computed are also indicated).
%
Note that the differences between the rotation angles assigned to each vector are consistent with the pairwise angles needed to rotate one vector to another, i.e., $\theta_i - \theta_j \approx \theta_{ij}$.
%
To compute these alignments, the task is formulated as an optimization problem, and then relaxed to an eigenproblem.
%
The top eigenvector(s) of a matrix (which contains information about all pairwise alignments) 
provides the (optimal) rotation angles assigned to each data point.
%
%Because the method considers {\em all} pairwise alignments, angular synchronization is often robust to noise.

\begin{figure}[t]
\includegraphics[width=8.4cm]{fig4}
\caption{(a) Set of vectors, each in a different orientation. The angles to align one vector to another are indicated. (b) The vectors from (a), each rotated so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in (a). (c) Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see \eqref{eq:dmaps_W}), so that close data points are connected by dark (``stronger'') edges. (d) The data in (c), colored by the first (non-trivial) eigenvector from the diffusion map computational procedure, $\phi_2$. The color intensity is monotonic with the arclength, thus parameterizing the curve.  }
%\label{fig:schematics}
\customlabel{fig:schematics}{4}
\customlabel{subfig:dmaps_edges}{\ref{fig:schematics}a}
\customlabel{subfig:dmaps_color}{\ref{fig:schematics}b}
\customlabel{subfig:synch1}{\ref{fig:schematics}c}
\customlabel{subfig:synch2}{\ref{fig:schematics}d}
\end{figure}

Figure \ref{subfig:1d_aligned_unordered} shows the concentration profiles from Figure~\ref{subfig:1d_unaligned_unordered}, now aligned using angular synchronization.
%
It is reassuring that the intensity peaks appear aligned, 
even though the algorithm uses no specific information about the existence or location of such peaks.
%
We note once more that, since angular synchronization only uses pairwise rotation information, 
the optimal rotations are recovered up to a global rotation.

@@

After geometrically registering the data (images or concentration profiles), we want to order them in time to reconstruct the developmental dynamics.
%
Our main assumption is that data variability
strongly correlates with time (see below).
%
We use diffusion maps to temporally order our data.
%
In general, diffusion maps is a dimensionality reduction technique, which aims to uncover low-dimensional, nonlinear structure in high-dimensional data sets \cite{coifman2005geometric}.
%
Figure~\ref{subfig:dmaps_edges} shows an example of a two-dimensional data set (the ``high-dimensional" data) which lies on a one-dimensional nonlinear curve (the ``low-dimensional" manifold).
%
Figure~\ref{subfig:dmaps_color} shows the same data, now colored by the leading diffusion map coordinate.
%
The color parameterizes the one-dimensional curve that our eye can visually detect.
%
Assume that our (one dimensional, or two-dimensional image) data also lies along a one-dimensional, nonlinear curve in a high-dimensional space, and that this curve is parameterized by time (e.g., in Figure~\ref{subfig:dmaps_color}, we assume that the data points are ordered from youngest to oldest along the curve).
%
Therefore, ordering data along this curve will effectively order data in time.
%
Because our data will be very high dimensional, we cannot do this by eye, and instead require an algorithm to automatically extract this one-dimensional structure.

To uncover a parameterization of this curve, we assume that data points which are close in the high-dimensional space should also be close {\em on the curve} (we would like images which look similar to be close together in our developmental trajectory).
%
%We therefore want to find a parameterization of the points, such that the points which are close in the original, high-dimensional space are {\em as close as possible} in the parameterization.
%
To emphasize this point, edges between the pairs of data points in Figure~\ref{subfig:dmaps_edges} are colored according to the distance between the data points, so that close data points have dark (``strong", ``meaningful") edges connecting them.
%
Visually, these dark edges delineate the one-dimensional curve; in contrast, the light edges are much less informative 
about the low-dimensional structure.
%
%We want points which are close in the high-dimensional space, i.e., connected by dark edges, to be close in our low-dimensional %parameterization.
%
This problem can be formulated as an optimization problem \cite{Belkin2003}, and the solution is given by the top (non-trivial) eigenvector of a Markov matrix which contains information about pairwise distances between the data points.
%
This procedure has been named diffusion maps \cite{coifman2005geometric}.
%
%The formulation extends from curves to manifolds (i.e., we can search for an $n$-dimensional, rather than a one-dimensional, parameterization of the data, such that data points which are close in the high-dimensional space are also close in the $n$-dimensional parameterization).

Given our set of registered signals, shown in Figure \ref{subfig:1d_aligned_unordered}, we can use diffusion maps to order them in time.
%
We use the Euclidean distance between the (discretized in $\mathbb{R}^{100}$) aligned concentration profiles (the rows of Figure \ref{subfig:1d_aligned_unordered}) as our pairwise distances between data points.
%
Figure \ref{subfig:1d_aligned_ordered} shows the data from Figure \ref{subfig:1d_aligned_unordered}, now sorted by the value of the first (non-trivial) eigenvector, $\phi_2$.
%
As expected, the intensity of the peak grows as a function of $\phi_2$ (since we constructed these data
to have a peak growing in time).
%
%Although diffusion maps can order the data, it cannot tell us about the speed of this trajectory.
%
%Furthermore, the direction of the trajectory is also unknown; determining which end of the trajectory corresponds to the beginning of development is a task that needs to be done by hand.

\subsection{Vector diffusion maps for simultaneous registration and ordering}

%We are often interested in both registering and ordering our data.
%
%Our proposed alignment methodology, angular synchronization, utilizes the eigendecompostion of the matrix $H \in \mathbb{R}^{md \times md}$, and our proposed ordering algorithm, diffusion maps, utilizes the eigendecomposition of the matrix $A \in \mathbb{R}^{m \times m}$.
%
Angular synchronization (for registration) can be combined with diffusion maps (for temporal ordering) in a single eigencomputation that allows us to {\em simultaneously} recover the optimal alignments {\em and} the right temporal ordering.
%
This technique is called vector diffusion maps \cite{singer2012vector}.
%
It both reduces the number of required eigencomputations, and is much more robust to noise, as mistakes in alignment are also considered when ordering the data.
%
Figure \ref{subfig:1d_aligned_ordered_vdm} shows the result of simultaneously registering and ordering the data in Figure \ref{subfig:1d_unaligned_unordered} using vector diffusion maps.
%
The data has been ordered by $\psi_{1, 3}$, the first (non-trivial) vector diffusion maps coordinate (see {\it Materials and Methods}).
